\begin{acknowledgements}
  We thank Prof.\ Pedro Ferreira and Dr.\ Charles Lawrence for very
  useful suggestions, comments and discussions.  We also thank the
  entire \Planck\ and \WMAP\ teams for invaluable support and
  discussions, and for their dedicated efforts through several decades
  without which this work would not be possible. This research has
  partially been carried out within the master- and PhD-level course
  ``AST9240 --- Cosmological component separation'' at the University
  of Oslo, and support for this has been provided by the Research
  Council of Norway through grant agreement no.\ 274990. The
  BeyondPlanck Collaboration has received funding from the European
  Union’s Horizon 2020 research and innovation programme under grant
  agreement numbers 776282, 772253, and 819478. In addition, the
  collaboration acknowledges support from ESA; ASI, CNR, and INAF
  (Italy); NASA and DoE (USA); Tekes, AoF, and CSC (Finland); RCN
  (Norway); ERC and PRACE (EU).  L.\,T.\,Hergt was supported by a UBC
  Killam Postdoctoral Research Fellowship.  The work of Tamaki
  Murokoshi was supported by MEXT KAKENHI Grant Number 18H05539 and
  the Graduate Program on Physics for the Universe (GP-PU), Tohoku
  University.  Part of the work by F.\, Rahman was carried out using
  the \texttt{Nova} and \texttt{Coolstar} clusters at the Indian
  Institute of Astrophysics, Bangalore.  The work of
  K. S. F. Fornazier, G. A. Hoerning, A. Marins and F. B. Abdalla was
  developed in the Brazilian cluster SDumont and was supported by
  S\~ao Paulo Research Foundation (FAPESP) grant 2014/07885-0,
  National Council for Scientific and Technological (CNPQ),
  Coordination for the Improvement of Higher Education Personnel
  (CAPES) and University of S\~ao Paulo (USP).  Some of the results in
  this paper have been derived using the \texttt{healpy}
  \citep{zonca2019} and \texttt{HEALPix} \citep{gorski2005} packages.
  This work made use of Astropy\footnote{http://www.astropy.org}
  \citep{astropy:2013, astropy:2018, astropy:2022}.
\end{acknowledgements}
